---
title: "Ebola facilities and beds, DRC 2026"
subtitle: "Bundibugyo virus disease outbreak: where patients could be treated, and how full it was"
date: last-modified
date-format: "D MMMM YYYY"
---

```{r setup}
source("_common.R")

drc <- capacity[country %in% c("Democratic Republic of the Congo", "") &
    (preferred)]

# Known open by a date: the latest date by which the reports show a site open
# or holding patients. A lower bound on how many sites were open on any day.
open_by <- opening[site_kind %in% c(PATIENT_KINDS, "hospital_isolation") &
    !is.na(opened_by)]
closed <- facilities[status_latest == "closed", facility_id]

days <- seq(report_first, report_last, by = "day")
open_curve <- open_by[, .(date = days, n = vapply(days,
    \(d) sum(opened_by <= d), integer(1))), by = site_kind]
open_curve[, kind := kind_label(site_kind)]

latest_n <- open_by[, .N, by = site_kind]
n_cte <- latest_n[site_kind == "treatment_centre", N]
n_ct <- latest_n[site_kind == "transit_centre", N]
n_ci <- latest_n[site_kind == "isolation_centre", N]

beds_nat <- drc[indicator == "beds_capacity" & level == "national"][
    order(as_of_date)]
occ_nat <- drc[indicator == "bed_occupancy_pct" & level == "national"][
    order(as_of_date)]
last_beds <- beds_nat[.N]
last_occ <- occ_nat[.N]
```

::: {.note}
This page summarises a dataset that is still under review. Every figure is
read by a language model from public situation reports and checked word for
word against its source, but most rows have not yet been read by a person.
Facility counts are an upper bound: where two names might be one site they are
counted as two until someone who knows the site says otherwise.
[Data and methods](methods.qmd) has the detail.
:::

## At a glance

Reports from `r fmt_date(report_first)` to `r fmt_date(report_last)` 2026.

| | |
|---|---|
| Treatment centres (CTE) reported open | `r n_cte` |
| Transit centres (CT) reported open | `r n_ct` |
| Isolation centres (CI) reported open | `r n_ci` |
| Active beds, `r fmt_date(last_beds$as_of_date)` (WHO AFRO) | `r fmt_n(last_beds$value)` |
| Bed occupancy, `r fmt_date(last_occ$as_of_date)` (WHO AFRO) | `r sprintf("%.0f%%", last_occ$value)` |

A site counts as open once a report shows it open or holding patients.
`r sum(open_by$facility_id %in% closed)` of these are later reported closed
and are still counted here.

## Sites open over time

```{r fig-open}
#| fig-cap: "Sites reported open or holding patients, by the date a report first showed it. Each line can only rise: a closure is not subtracted. The true opening of most sites is earlier than the line shows, because the first report to mention a site usually finds it already working."
#| fig-alt: "Step lines of cumulative sites open by kind, rising from mid-May to mid-September 2026. Hospital isolation and treatment centres grow most."
ends <- open_curve[date == max(date)]
ggplot(open_curve, aes(date, n, group = kind)) +
    geom_step(aes(colour = kind == KINDS["treatment_centre"]),
        linewidth = 0.7) +
    geom_text(data = ends, aes(label = paste0(kind, "  ", n)),
        hjust = 0, nudge_x = 2, size = 3.2, colour = INK) +
    scale_colour_manual(values = c(`TRUE` = ACCENT, `FALSE` = INK),
        guide = "none") +
    scale_x_outbreak() +
    coord_cartesian(clip = "off") +
    labs(x = NULL, y = "Sites open (cumulative)") +
    theme(plot.margin = margin(5, 150, 5, 5))
```

## Beds and patients

```{r fig-beds}
#| fig-cap: "National active bed capacity and patients in isolation. WHO AFRO figures are weekly; INSP figures are daily and stop in early August, when the bed tables leave the situation reports. Lines break where no figure was published for more than ten days."
#| fig-alt: "Line chart of bed capacity rising from about 650 in late June to about 1500 in September, with patients in isolation between 700 and 900."
beds_series <- drc[level == "national" &
    indicator %in% c("beds_capacity", "patients_in_isolation") &
    period == "point"]
beds_series[, what := fifelse(indicator == "beds_capacity",
    "Active beds", "Patients in isolation")]
beds_series[, seg := gap_group(as_of_date), by = .(source, what)]
ggplot(beds_series, aes(as_of_date, value, colour = source,
    linetype = what, group = interaction(source, what, seg))) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1) +
    scale_colour_manual(values = SOURCE_COLOURS, labels = SOURCE_LABELS) +
    scale_y_continuous(labels = scales::label_comma(),
        limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    scale_x_outbreak() +
    labs(x = NULL, y = NULL)
```

```{r fig-occupancy}
#| fig-cap: "Bed occupancy by province, where a report gives it. Above the line at 100% a province held more patients than it had beds. Figures are as printed, not recomputed."
#| fig-alt: "Small multiple dot-and-line charts of bed occupancy by province. Nord-Kivu is repeatedly above 100 percent; Ituri falls from about 90 to about 45 percent."
#| fig-height: 5
occ <- drc[indicator == "bed_occupancy_pct" &
    level %in% c("province", "national")]
occ[, panel := fifelse(level == "national", "All provinces", place)]
occ[, panel := factor(panel, levels = c("All provinces", PROVINCES))]
occ[, seg := gap_group(as_of_date), by = .(source, panel)]
ggplot(occ[!is.na(panel)], aes(as_of_date, value, colour = source,
    group = interaction(source, seg))) +
    geom_hline(yintercept = 100, colour = GREY, linewidth = 0.3) +
    geom_line(linewidth = 0.5) +
    geom_point(size = 1.1) +
    facet_wrap(~panel, ncol = 3, drop = TRUE) +
    scale_colour_manual(values = SOURCE_COLOURS, labels = SOURCE_LABELS) +
    scale_y_continuous(limits = c(0, NA),
        expand = expansion(mult = c(0, 0.05))) +
    scale_x_outbreak() +
    labs(x = NULL, y = "Occupancy (%)")
```

<!-- UNCERTAIN: occupancy above 100% in Nord-Kivu is printed by WHO AFRO in
several weeks. It is plausible (patients on mattresses, transit overflow) but
not explained in the reports; it is shown as printed. -->

## Treatment centres

```{r tab-cte}
cte <- opening[site_kind == "treatment_centre"]
cte <- merge(cte, facilities[, .(facility_id, status_latest_date)],
    by = "facility_id", all.x = TRUE)
cte[, status := fcase(
    status_latest == "operating", "Holding patients",
    status_latest == "strained", "Strained",
    status_latest == "incident", "Incident reported",
    status_latest == "under_construction", "Being built",
    status_latest == "planned", "Planned",
    status_latest == "opened", "Opened",
    status_latest == "closed", "Closed",
    status_latest == "expanded", "Expanded",
    default = "Named only")]
setorder(cte, province, health_zone, facility_name)
html_table(cte[, .(
    Province = fifelse(province == "", "not stated", province),
    `Health zone` = health_zone,
    Centre = facility_name,
    `Open by` = fifelse(is.na(opened_by), "", fmt_date(opened_by)),
    `Latest status` = paste0(status, ", ", fmt_date(status_latest_date)),
    Beds = fifelse(is.na(beds_latest), "", as.character(beds_latest))
)], align = c("l", "l", "l", "r", "l", "r"))
```

"Open by" is the date of the first report showing the centre open or holding
patients. Beds are the latest count a report gives for that centre; most
centres never have one, because the facility bed table stops appearing after
SitRep 090.

## Where this comes from

The facility list and opening dates are read from the Institut National de
Santé Publique (INSP) situation reports. The bed figures after early August
come from the World Health Organization Regional Office for Africa (WHO AFRO)
weekly situation reports, because the INSP bed tables stop. Every row in the
underlying data carries the sentence it came from.

Data: [epiforecasts/bvd-capacity](https://github.com/epiforecasts/bvd-capacity).
Please cite the INSP and WHO reports for any figure you use.
