---
title: "Data and methods"
subtitle: "What the tables hold, how far they can be trusted, and how to use them in a model"
date: last-modified
date-format: "D MMMM YYYY"
---

```{r setup}
source("_common.R")

decisions <- fread(data_path("..", "registry", "decisions.csv"))
rejected <- fread(data_path("..", "checks", "rejected_events.csv"))
conflicts_path <- data_path("..", "checks", "capacity_conflicts.csv")
conflicts <- if (file.exists(conflicts_path)) fread(conflicts_path) else NULL
open_sheets <- length(list.files(data_path("..", "checks", "decisions"),
    pattern = "\\.md$"))

drc <- capacity[country %in% c("Democratic Republic of the Congo", "")]
has_insp <- drc[source == "insp", .N] > 0
```

This page is for anyone using the data in an analysis. The
[overview](index.qmd) is the short version. The column definitions are in
[`dictionary.md`](https://github.com/epiforecasts/bvd-capacity/blob/main/dictionary.md).

## The tables

```{r tab-files}
files <- data.table(
    File = c("facility_events.csv", "facilities.csv", "facility_opening.csv",
        "capacity_indicators.csv"),
    `One row is` = c(
        "one thing one report says about one facility",
        "one facility, with dates derived from its events",
        "one facility, with the interval its opening falls in",
        "one figure: an indicator, a place, a date, a source"),
    Rows = fmt_n(c(nrow(events), nrow(facilities), nrow(opening),
        nrow(capacity))),
    `Use it for` = c(
        "anything that should respect what was known when",
        "a register of sites; a join key",
        "opening dates, with their censoring",
        "system bed capacity and occupancy over time")
)
html_table(files, align = c("l", "l", "r", "l"))
```

Every row in the first and last tables carries the sentence or table row it
came from, in `evidence_quote`, and the report it came from. Read the quote
before relying on a row.

Loading, in R:

```r
library(data.table)
base <- "https://raw.githubusercontent.com/epiforecasts/bvd-capacity/main/data/"
events <- fread(paste0(base, "facility_events.csv"))
opening <- fread(paste0(base, "facility_opening.csv"))
capacity <- fread(paste0(base, "capacity_indicators.csv"))

# The national bed series a model should use: DRC only, one figure a key
beds <- capacity[country == "Democratic Republic of the Congo" &
    indicator == "beds_capacity" & level == "national" & preferred]
```

Pin a commit rather than `main` for anything you publish: ids change as
naming decisions land.

## Reports read

```{r fig-coverage}
#| fig-cap: "Situation reports read, by week. Seven SitRep numbers were never published; a facility first mentioned just after a gap may have been open earlier."
#| fig-alt: "Bar chart of reports per week, mostly six or seven a week from late May to mid-September."
#| fig-height: 2.5
reps <- unique(events[, .(sitrep, report_date)])
reps[, week := as.IDate(cut(as.Date(report_date), "week"))]
ggplot(reps[, .N, week], aes(week, N)) +
    geom_col(fill = INK, width = 5) +
    scale_x_outbreak() +
    labs(x = NULL, y = "Reports")
```

`r uniqueN(events$sitrep)` reports, SitReps 001 to 122, from
`r fmt_date(report_first)` to `r fmt_date(report_last)`. Missing numbers: 003,
029, 043, 045, 063, 075, 076. `006_v2` is a reissue kept as its own report.

## From sentences to facilities

```{r tab-resolution}
res <- events[, .N, by = name_status][order(-N)]
res[, what := fcase(
    name_status == "named", "named, resolved to a facility id",
    name_status == "ambiguous", "a name that could be several sites",
    name_status == "unnamed", "a site described but not named")]
html_table(res[, .(Events = fmt_n(N), `What the report gave` = what)],
    align = c("r", "l"))
```

A language model (`gemini-3.1-pro`, low thinking, one call a report) reads
each report and returns events in a fixed vocabulary, each with a quote. A
quote that is not a character-for-character span of the report is dropped;
`r nrow(rejected)` events are dropped this way and listed in
`checks/rejected_events.csv`. Names are then keyed and matched to a
hand-maintained vocabulary of spellings.

Code cannot tell a misspelling from a second centre at the same hospital, so
where two names might be one site they stay apart and are flagged. A person
settles each question and the answer is recorded with who decided and why.

```{r tab-decisions}
dec <- decisions[, .N, by = .(verdict, decided_by)]
dec[, verdict := fcase(verdict == "same", "merged into one facility",
    verdict == "apart", "kept apart, deliberately",
    verdict == "unsure", "looked at, left open")]
html_table(dec[, .(Decision = verdict, `Decided by` = decided_by,
    Questions = N)], align = c("l", "l", "r"))
```

`claude-code` decisions were made by the assistant that built the pipeline
from GRID3 names and spelling evidence, and nobody who knows the sites has
checked them. `r open_sheets - 1` questions have evidence sheets in
`checks/decisions/` and are unanswered. Until they are, facility counts are an
upper bound: the error runs towards counting one site twice, never towards
merging two.

## Opening dates

The reports almost never give an opening date. An opening is inferred from
two kinds of report: one showing a site being built or planned (it had not
opened by then) and one showing it open or holding patients (it had opened
by then).

```{r tab-censoring}
cens <- opening[site_kind %in% c(PATIENT_KINDS, "hospital_isolation")]
cens[, censoring := sub(":.*", "", censoring)]
tab <- dcast(cens, site_kind ~ censoring, value.var = "facility_id",
    fun.aggregate = length)
tab[, site_kind := as.character(kind_label(site_kind))]
setnames(tab, "site_kind", "Kind of site")
html_table(tab)
```

`bounded` is a site seen building and later seen open, so both ends of its
opening are known. `left censored` is first seen already open: the opening
is at or before that date and nothing earlier is known. `right censored` was
seen building and never seen open. For most sites the only honest statement
is "open by this date".

```{r fig-cte-timeline}
#| fig-cap: "Treatment centres: the interval each opening falls in. A segment runs from the last report showing the centre not yet open to the first showing it open. A lone dot is left censored: open by then, earlier unknown. An open circle was never seen open, and sits at the last report showing it not yet open."
#| fig-alt: "Timeline of about seventy treatment centres ordered by opening. Most are single dots; thirteen have short segments of a few days."
#| fig-height: 9
tl <- opening[site_kind == "treatment_centre"]
tl[, y := fifelse(is.na(opened_by), opened_after, opened_by)]
tl <- tl[!is.na(y)]
setorder(tl, y)
id_order <- rev(tl$facility_id)
ggplot(tl, aes(y = facility_id)) +
    geom_segment(data = tl[!is.na(opened_after) & !is.na(opened_by)],
        aes(x = opened_after, xend = opened_by, yend = facility_id),
        colour = ACCENT, linewidth = 0.8) +
    geom_point(data = tl[!is.na(opened_by)], aes(x = opened_by),
        colour = INK, size = 1.2) +
    geom_point(data = tl[is.na(opened_by)], aes(x = opened_after),
        shape = 1, colour = GREY, size = 1.4) +
    scale_y_discrete(limits = id_order) +
    scale_x_outbreak() +
    labs(x = NULL, y = NULL) +
    theme(axis.text.y = element_text(size = 6.5),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(colour = RULE, linewidth = 0.3))
```

A date in these tables is the date a report said something, not the date it
happened. For a model of when capacity came online, treat `opened_by` as an
upper bound on the opening and model the delay from opening to first mention.

## The capacity series

`capacity_indicators.csv` has a fixed vocabulary of ten indicators. The INSP
situation reports carry bed tables until early August; WHO AFRO's weekly
reports carry national and province figures from late June. Where they
overlap the two can be checked against each other.

```{r fig-cap-coverage}
#| fig-cap: "Which indicators each source gives, when, at national or province level. One mark a figure."
#| fig-alt: "Dot matrix of indicators against date, coloured by source."
#| fig-height: 4
cov <- drc[level %in% c("national", "province")]
ggplot(cov, aes(as_of_date, indicator, colour = source)) +
    geom_point(shape = 124, size = 3) +
    scale_y_discrete(limits = rev) +
    facet_wrap(~level, ncol = 1) +
    scale_colour_manual(values = SOURCE_COLOURS, labels = SOURCE_LABELS) +
    scale_x_outbreak() +
    labs(x = NULL, y = NULL)
```

### When a report gives two figures

A report sometimes gives two different figures for the same indicator, place
and date. Every figure stays in the table. The column `preferred` marks the
one a series should use, chosen by these rules in order, decided by Kath
Sherratt on 26 September 2026 and recorded in
`registry/capacity_decisions.csv`:

1. "Increased from A to B": B is current; A belongs to the week before.
2. Where the report states a numerator and denominator, the occupancy
   consistent with them is preferred over a printed percentage that
   contradicts them. Where no beds-occupied figure is printed, patients in
   isolation stands in as the numerator.
3. Otherwise the headline figure box beats a narrative sentence.

```{r tab-conflicts, eval = !is.null(conflicts)}
conflicts[, how := fcase(
    startsWith(rule, "rule1"), "1: from A to B",
    startsWith(rule, "rule2"), "2: numerator and denominator",
    startsWith(rule, "rule3"), "3: headline box",
    startsWith(rule, "decision"), "read by hand",
    default = "not settled")]
cf <- conflicts[, .(Figures = .N), by = .(Indicator = indicator,
    `Settled by` = how)]
html_table(cf[order(Indicator)], align = c("l", "l", "r"))
```

<!-- UNCERTAIN: rule 2 treats patients in isolation as beds occupied when
no occupied count is printed (afro-12). Isolation can include patients in
transit centres not counted in bed capacity. See DEVLOG item 10. -->

A key with no preferred figure is one the rules could not settle; it is
listed in `checks/capacity_conflicts.csv` and left out of any series.

```{r overlap-intro, eval = has_insp, results = "asis"}
cat("### INSP against WHO AFRO\n\n")
```

```{r fig-overlap, eval = has_insp}
#| fig-cap: "Patients in isolation and bed occupancy, national, where both sources give a figure. INSP is daily, WHO AFRO weekly."
#| fig-alt: "Two panels comparing INSP and WHO AFRO national figures over the weeks both report."
#| fig-height: 4
ov <- drc[preferred & level == "national" &
    indicator %in% c("patients_in_isolation", "bed_occupancy_pct",
        "beds_capacity")]
span <- ov[, .(from = min(as_of_date), to = max(as_of_date)), by = source]
lo <- max(span$from) - 14
hi <- min(span$to) + 14
ov <- ov[between(as_of_date, lo, hi)]
ov[, seg := gap_group(as_of_date), by = .(source, indicator)]
ggplot(ov, aes(as_of_date, value, colour = source,
    group = interaction(source, seg))) +
    geom_line(linewidth = 0.5) +
    geom_point(size = 1) +
    facet_wrap(~indicator, scales = "free_y", ncol = 1) +
    scale_colour_manual(values = SOURCE_COLOURS, labels = SOURCE_LABELS) +
    scale_x_date(date_labels = "%e %b") +
    labs(x = NULL, y = NULL)
```

## Before using this in a model

- Filter on `country`. The WHO AFRO rows include Uganda, and
  `r drc[country == "", .N]` rows have no country.
- Use `preferred`. Without it a national series double-counts dates where a
  report printed two figures.
- Occupancy above 100 % is printed by WHO AFRO for Nord-Kivu in several weeks.
  It is kept as printed; the reports do not say what it counts.
- Absence is not zero. A figure a report did not print is missing, never 0.
- Facility counts are an upper bound until the open naming questions are
  settled.
- Facility bed counts are sparse: `r facilities[!is.na(beds_latest), .N]`
  facilities have one, because the INSP facility bed table stops after SitRep
  090.
- `r events[name_status != "named", .N]` events describe a site without a
  usable name. They are in `facility_events.csv` with an empty `facility_id`
  and absent from every per-facility table.

## Sources and citation

Institut National de Santé Publique, DRC (2026), situation reports on the
17th Ebola virus disease epidemic, <https://insp.cd/ebola-17eme-epidemie/>,
via the corpus in [bvd-sitreps](https://github.com/epiforecasts/bvd-sitreps).
WHO Regional Office for Africa weekly situation reports, CC BY-NC-SA 3.0 IGO.
GRID3 COD Health Facilities v8.0, CC BY 4.0,
[doi:10.7916/f1ft-y872](https://doi.org/10.7916/f1ft-y872).

Cite the report a figure came from. If you relied on the facility identities
or opening intervals, also cite
[bvd-capacity](https://github.com/epiforecasts/bvd-capacity).
